---
title: "libvroom Benchmarks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{libvroom Benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(ggplot2)
library(forcats)
library(dplyr)
library(tidyr)
library(fs)

pretty_sec <- function(x) {
  x[!is.na(x)] <- prettyunits::pretty_sec(x[!is.na(x)])
  x
}

read_benchmark <- function(file, desc) {
  vroom::vroom(file, col_types = c("ccccddddd")) |>
    filter(op != "setup") |>
    mutate(
      reading_package_label = case_when(
        reading_package == "vroom_cran" ~ "vroom (CRAN)",
        reading_package == "vroom_arrow" ~ "vroom (Arrow)",
        reading_package == "data.table" ~ "data.table",
        TRUE ~ reading_package
      ),
      label = fct_reorder(
        paste0(reading_package_label, "\n", manip_package),
        case_when(type == "real" ~ time, TRUE ~ 0),
        sum
      ),
      op = factor(op, desc)
    )
}

generate_subtitle <- function(data) {
  rows <- scales::comma(data$rows[[1]])
  cols <- scales::comma(data$cols[[1]])
  size <- fs_bytes(data$size[[1]])
  glue::glue("{rows} x {cols} - {size}B")
}

plot_benchmark <- function(data, title) {

  subtitle <- generate_subtitle(data)
  data <- data |>
    filter(type == "real")

  p1 <- data |>
    ggplot() +
    geom_bar(aes(x = label, y = time, fill = op, group = label), stat = "identity") +
    scale_fill_brewer(type = "qual", palette = "Set2") +
    scale_y_continuous(labels = scales::number_format(suffix = "s")) +
    coord_flip() +
    labs(title = title, subtitle = subtitle, x = NULL, y = NULL, fill = NULL) +
    theme(legend.position = "bottom")

  p2 <- data |>
    group_by(label) |>
    summarise(max_memory = max(max_memory)) |>
    ggplot() +
    geom_bar(aes(x = label, y = max_memory / (1024 * 1024)), stat = "identity") +
    scale_y_continuous(labels = scales::number_format(suffix = "Mb")) +
    coord_flip() +
    labs(title = "Maximum memory usage", x = NULL, y = NULL) +
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

  library(patchwork)
  p1 + p2 + plot_layout(widths = c(2, 1))
}

make_table <- function(data) {
  data |>
    filter(type == "real") |>
    select(-label, -size, -type, -rows, -cols) |>
    mutate(reading_package = reading_package_label) |>
    select(-reading_package_label) |>
    spread(op, time) |>
    mutate(
      total = read + materialize + print + head + tail + sample + filter + aggregate,
      max_memory = as.character(bench::as_bench_bytes(max_memory))
    ) |>
    arrange(desc(total)) |>
    mutate_if(is.numeric, pretty_sec) |>
    select(reading_package, manip_package, max_memory, everything()) |>
    rename(
      "reading\npackage" = reading_package,
      "manipulating\npackage" = manip_package,
      memory = max_memory
    ) |>
    knitr::kable(digits = 2, align = "r", format = "html")
}

desc <- c("setup", "read", "materialize", "print", "head", "tail", "sample", "filter", "aggregate")
```

These benchmarks compare `vroom` with its new libvroom SIMD-accelerated backend
against CRAN vroom, `arrow`, and `data.table` across several data reading and
manipulation scenarios. Each benchmark reads a file, then performs a series of
common operations: printing, taking head/tail, sampling, filtering, and
aggregation.

The libvroom backend uses portable SIMD instructions (via Google Highway) for
CSV parsing and outputs data in Arrow columnar format for efficient R
integration.

# Reading delimited files

The following benchmarks measure reading delimited files and performing common
data manipulation tasks. A "materialize" step is included after reading to
convert lazy representations into fully realized R vectors, providing a fair
comparison across packages.

## Mixed-type data

This dataset contains a mix of character, integer, and double columns,
representing a typical real-world CSV file.

```{r}
#| warning: false
#| echo: false
#| message: false
#| fig-height: 8
#| fig-width: 10
#| fig-alt: "Horizontal bar chart comparing time and memory usage across different R packages (vroom, vroom CRAN, arrow, data.table) for reading and analyzing mixed-type data. The chart shows operation breakdowns including read, materialize, print, head, tail, sample, filter, and aggregate operations. Consult the associated table for detailed timing and memory usage statistics."
mixed <- read_benchmark(path_package("vroom", "bench2", "mixed.tsv"), desc)

plot_benchmark(mixed, "Time to analyze mixed-type data")

make_table(mixed)
```

## All-numeric data

All-numeric data is a challenging scenario for lazy readers because the index
takes about as much memory as the parsed data. Numeric values also parse
quickly in parallel, so there is less room for improvement from deferred
evaluation. This benchmark highlights raw parsing throughput.

```{r}
#| warning: false
#| message: false
#| echo: false
#| fig-height: 8
#| fig-width: 10
#| fig-alt: "Horizontal bar chart comparing time and memory usage for reading and analyzing all-numeric data across different R packages (vroom, vroom CRAN, arrow, data.table). Shows performance breakdown by operation type."
all_num <- read_benchmark(path_package("vroom", "bench2", "all_numeric.tsv"), desc)

plot_benchmark(all_num, "Time to analyze all-numeric data")

make_table(all_num)
```

## All-character data

All-character data benefits most from lazy evaluation, as character parsing and
memory allocation are the most expensive operations. Packages that defer
materialization can skip parsing entirely for columns that are never accessed.

```{r}
#| warning: false
#| message: false
#| echo: false
#| fig-height: 8
#| fig-width: 10
#| fig-alt: "Horizontal bar chart comparing time and memory usage for reading and analyzing all-character data across different R packages (vroom, vroom CRAN, arrow, data.table). Shows performance breakdown by operation type."
all_chr <- read_benchmark(path_package("vroom", "bench2", "all_character.tsv"), desc)

plot_benchmark(all_chr, "Time to analyze all-character data")

make_table(all_chr)
```

## Session and package information

```{r}
#| echo: false
#| warning: false
#| message: false
si <- vroom::vroom(path_package("vroom", "bench2", "session_info.tsv"))
class(si) <- c("packages_info", "data.frame")
select(as.data.frame(si), package, version = ondiskversion, date, source) |>
  knitr::kable()
```
